---
title: "Music and Chronic Pain"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Fill in abstract... Write this last, after finishing methods, results, and discussion. Summarize the overall study question, approach, results, and conclusion in a short paragraph.

# BACKGROUND


# STUDY QUESTION and HYPOTHESIS

## Questions  

Does music decrease the sensation of pain in individuals with chronic pain?

## Hypothesis 

Listening to music decreases the sensation of pain.

## Prediction

Participants who listen to music will experience a lower sensation of pain.

# METHODS

Fill in here... Summarize the procedures and analyses you use in this report. In this section, describe what you did and why. Don’t just restate the code — explain the logic of each analysis in plain language. Keep each subsection short (2–4 sentences).

#STATISTICAL ANALYSIS

To visually compare the control (pink noise) against the treatment group (music), a table of values and barplot were contstructed measuring the pain sensation levels between the two groups.

To statistically test the data, we will perform an ANOVA test to compare the means.
###*Figure 1*
```{r first-analysis, warning=FALSE}
#simply showing the data captured in the table with values.
read.csv("musicdataset.csv")
```


# DISCUSSION

Fill in here... For each analysis, summarize what you found and interpret the results. Say whether they support or contradict your hypothesis.

# CONCLUSION

Fill in here... State the overall answer to your research question, based on all analyses. Mention whether the evidence supports your hypothesis.

# REFERENCES     
Fill in here... List all sources you cited in your background and throughout the report. Use a consistent style.a conclusion you can draw from your analysis.



