---
title: "Music and Chronic Pain"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT

Fill in abstract... Write this last, after finishing methods, results, and discussion. Summarize the overall study question, approach, results, and conclusion in a short paragraph.


# BACKGROUND

Music has been a deep and important part of human culture and expression for millennia. Yet, not much is understood about how music affects the human body physiologically. As research develops, music has been shown to provide a myriad of benefits to human quality of life. Recently, S. Aalbers and others found that when combined with treatment as usual, music had a significant positive effect on patients with depression (Aalbers, 2017).

This study explores how music affects the pain sensation of individuals diagnosed with chronic pain. The results will provide exciting insights to a potential link between music and pain reduction.

# STUDY QUESTION and HYPOTHESIS

## Questions  

Does music affect the sensation of pain in individuals with chronic pain?

## Hypothesis 

Listening to music decreases the sensation of pain in individuals with chronic pain.

## Prediction

Participants who listen to music will report a lower sensation of pain.

# METHODS

Fill in here... Summarize the procedures and analyses you use in this report. In this section, describe what you did and why. Don’t just restate the code — explain the logic of each analysis in plain language. Keep each subsection short (2–4 sentences).

#STATISTICAL ANALYSIS

To visually compare the control (pink noise) against the treatment group (music), a table of values and barplot were contstructed measuring the pain sensation levels between the two groups.

To statistically test the data, we will perform an ANOVA test to compare the means.
###*Figure 1*
```{r first-analysis, warning=FALSE}
#simply showing the data captured in the table with values.
read.csv("musicdataset.csv")
```

```{r boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
data_long <- data %>%
  pivot_longer(cols = c(5, 6,7,8), names_to = "treatment", values_to = "value")

#plot
    ggplot(data_long, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_brewer(palette="Spectral") +
    theme(axis.text.x = element_text(size=15)) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    ylab("Pain Level") +
    xlab("")
```

``` {r}
ggplot(
  data_long %>%
    mutate(
      treatment = recode(treatment,
                         pic1 = "Before Control",
                         pic2 = "After Control",
                         pim1 = "Before Music",
                         pim2 = "After Music"),
      # optional: control order on x-axis
      treatment = factor(treatment, levels = c("Before Control", "After Control", "Before Music", "After Music"))
    ),
  aes(x = treatment, y = value, fill = treatment)
) +
  geom_boxplot(alpha = 0.6) +
  scale_y_continuous(breaks = seq(0, 10, 1)) +
  coord_cartesian(ylim = c(0, 10)) +
  scale_fill_brewer(palette = "Spectral") +
  theme_classic() +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 12, face = "bold"),
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 16, face = "bold", margin = margin(t = 10)),
    axis.title.y = element_text(size = 16, face = "bold", margin = margin(r = 10))
  ) +
  xlab("Pain Intensity") +
  ylab("Pain Level")
```

###*Figure 2*
``` {r code pic chunks}

library(ggplot2)
library(tidyr)
library(dplyr)

data <- read.csv("musicdataset.csv")

# Pivot pic1 and pic2 into long format
data_long <- data %>%
  pivot_longer(cols = c(pic1, pic2), names_to = "treatment", values_to = "value")

# Make boxplot
ggplot(data_long, aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  theme_classic() +
  scale_fill_brewer(palette = "Set2") +
  ylab("Pain Level") +
  xlab("") +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )

```

pim 1 and 2 code chunk 

``` {r code chunk pim}
library(ggplot2)
library(tidyr)
library(dplyr)

data <- read.csv("musicdataset.csv")

# Pivot pim1 and pim2 into long format
data_long_pim <- data %>%
  pivot_longer(cols = c(pim1, pim2), names_to = "treatment", values_to = "value")

# Make boxplot
ggplot(data_long_pim, aes(x = treatment, y = value, fill = treatment)) +
  geom_boxplot(alpha = 0.7, width = 0.6) +
  theme_classic() +
  scale_fill_brewer(palette = "Set2") +
  ylab("Pain Level") +
  xlab("") +
  theme(
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16, face = "bold"),
    legend.position = "none"
  )
```


##Statistical Analysis
Still working on this part...had to remove so file could knit.

# DISCUSSION

Fill in here... For each analysis, summarize what you found and interpret the results. Say whether they support or contradict your hypothesis.

# CONCLUSION

Fill in here... State the overall answer to your research question, based on all analyses. Mention whether the evidence supports your hypothesis.

# REFERENCES     
Aalbers, S., Fusar‐Poli, L., Freeman, R. E., Spreen, M., Ket, J. C., Vink, A. C., ... & Gold, C. (2017). Music therapy for depression. Cochrane database of systematic reviews, (11).

Fill in here... List all sources you cited in your background and throughout the report. Use a consistent style.a conclusion you can draw from your analysis.



