---
title: "Music and Chronic Pain"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT
This study investigates whether listening to music can reduce the sensation of pain in individuals diagnosed with chronic pain due to fibromyalgia. Participants rated their pain before and after listening to either five minutes of self-selected pleasant music or a pink-noise control stimulus. Boxplot visualizations and paired t-tests were used to compare pre- and post-treatment pain levels across both conditions and test for statistical significance. Results showed a statistically significant decrease in pain following the music condition, while the control condition produced no meaningful reduction and even suggested a slight increase in reported pain. These findings support the hypothesis that listening to music decreases the sensation of pain in individuals with chronic pain from fibromyalgia.

# BACKGROUND

Music has been a deep and important part of human culture and expression for millennia. Yet, much is still to be understood about how music affects the human body physiologically. As research develops, music has been shown to provide a myriad of benefits to human quality of life. Recently, Aalbers *et al.* (2017) found that when combined with treatment as usual, music had a significant positive effect on patients with depression.

Fibromyalgia is a chronic medical condition classified as a central pain syndrome. It brings along a host of unpleasant symptoms including irritable bowel syndrome, temporomandibular disorder, chronic low back pain, and certain other chronic pain disorders (Clauw, 2009).

This study explores how music affects the pain sensation of individuals diagnosed with chronic pain from fibromyalgia. It addresses a significant need in our society since many traditional medicine and treatments have been largely deemed unsatisfactory by patients. The results provide exciting insights into a potential link between music and pain reduction.

# STUDY QUESTION and HYPOTHESIS

## Questions

Does music affect the sensation of pain in individuals with chronic pain from fibromyalgia?

## Hypothesis 

Listening to music decreases the sensation of pain in individuals with chronic pain from fibromyalgia.

## Prediction

Participants of the study will report a lower sensation of pain after listening to music.

# METHODS

The dataset was provided by Garza-Villarreal *et al.* (2019). The study collected the pain data of 20 individuals diagnosed with fibromyalgia. Each participant underwent a control and treatment test. Pain levels were self-reported on a scale of 1-10 before and after each test. In the control test, participants listened to pink noise for 5 minutes. In the treatment test, participants listened to music chosen by themselves. By requirement, the chosen music had a tempo less than 120 bpm and was deemed familiar and pleasant by the participant. No other information about the music selection or patient pool was provided in the dataset.

Boxplots were constructed using R to visualize data. A t-test was then performed to analyze the statistical significance of the mean differences between before and after groups seen in the boxplot.

## *Figure 1* - Boxplot of Before and After Control and Treatment
Visualization of four data groups (Before Control, After Control, Before Music, After Music) was presented via boxplots. All 4 plots are presented first, then control and treatment groups are separated for clarity.

```{r total boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#control treatment
data_total <- data %>%
  select(1, 5, 6, 7, 8) %>%
  pivot_longer(
    cols = c(2, 3, 4, 5),
    names_to = "treatment",
    values_to = "value" )


#plot control
    ggplot(data_total, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_manual(values = c("pic1" ="#B16C02" , "pic2" = "#FFB501", "pim1" ="#003F80" , "pim2" = "#1E88E5")) +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
      scale_x_discrete(labels = c("pic1" = "Before Control", "pic2" = "After Control", "pim1" = "Before Music", "pim2" = "After Music"))
    
```

## *Figure 2* - Control Boxplots: Before and After
```{r control boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#control treatment
data_control <- data %>%
  select(1, 5, 6) %>%
  pivot_longer(
    cols = c(2, 3),
    names_to = "treatment",
    values_to = "value" )


#plot control
    ggplot(data_control, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_manual(values = c("pic1" ="#B16C02" , "pic2" = "#FFB501")) +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
      scale_x_discrete(labels = c("pic1" = "Before Control", "pic2" = "After Control"))
```


## *Figure 3* - Treatment Boxplots: Before Music and After Music
```{r music boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#music treatment
data_music <- data %>%
  select(1, 7, 8) %>%
  pivot_longer(
    cols = c(2, 3),
    names_to = "treatment",
    values_to = "value")

#plot music
    ggplot(data_music, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
    scale_fill_manual(values = c("pim1" ="#003F80" , "pim2" = "#1E88E5")) +
    theme_classic() +
    theme(legend.position="none") +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
    scale_x_discrete(labels = c("pim1" = "Before Music", "pim2" = "After Music"))
    
```



## Statistical Analysis
### *Figure 4* - T-Test of Control: Before and After
T-test was performed to analyze the difference in means before and after pink noise (control) and test for significance.
```{r t test control}
t.test(data$pic2, data$pic1, paired = TRUE)

```
### *Figure 5* - T-Test of Treatment: Before and After
T-test was performed to analyze the difference in means before and after music (treatment) and test for significance.
```{r t test music}
t.test(data$pim2, data$pim1, paired = TRUE)

```
# RESULTS
## Boxplot Figures
The boxplots visualize a the trends of the pain level reports before and after control and treatment groups. In the control group, the average and lower quartile of reported pain level increased in the after control group. The upper quartile did not change, and the entire range of reports remained constant (0-10). In the treatment group, the average and distribution (upper and lower quartiles) of reported pain level decreased in the after music group. Again, the entire range remained constant (0-10). The average and distribution of before control and after music groups appear identical.

## Statistical Analysis
### *Figure 4*
The control (pink noise) had a marginally significant effect (*p*-vaue: 0.07204) on pain sensation. The mean difference was 0.4705882, suggesting a positive effect.

### *Figure 5*
The treatment (music) had a significant effect (*p*-value:0.007141) on pain sensation. The mean difference was -1.117647, suggesting a negative effect.

# DISCUSSION
The results show statistical evidence suggesting music leads to a decrease in pain sensation. The control exhibited moderate significance for the pink noise to affect pain levels in the positive direction, meaning there may be evidence suggesting that pink noise increased pain. In this study, it is possible that untreated pain increased over time, adding more statistical significance to the decrease found in the treatment group. The null hypothesis that music does not affect pain sensation can be confidently rejected. Furthermore, there was a greater mean difference that was in the negative direction for music, reflecting a visible decrease in the average reported pain levels in patients.

The results were limited due to individual interpretation of pain. Each participant carried a distinct pain threshold and reporting behavior. These effects could be reduced by using a visual pain scale, such as the Wong-Baker FACES Scale, which provides a visual reference point to help patients determine pain levels on a scale that could be more universal. Results were also hindered due to the small sample size of only 20 patients. Larger, more diverse samples (including patients with pain not associated with fibromyalgia) would yield results with stronger statistical support and broader implications.

Further research should compare both different forms of “noise” (white, brown, etc.) and multiple music genres to other methods of pain reduction, such as medication or physical/talk therapy. Regardless, the evidence showing decreased pain from music is a step towards a more comfortable future in patients with this diagnosis, as they may find affordable and effective methods for pain relief.

# CONCLUSION
After testing the pain sensation of individuals with chronic pain due to fibromyalgia with and without music, it can be concluded with strong statistical significance (*p*-value: 0.007141, Mean Difference: -1.117647) that listening to familiar, pleasant music reduces the sensation of pain. Control conditions of exposure to pink noise exhibited an increase in reported pain sensation with moderate to low statistical strength (*p*-value: 0.07204, Mean Difference: 0.4705882). These findings suggest an exciting new method of pain relief for patients suffering from fibromyalgia-related chronic pain.

# REFERENCES     
Aalbers, S., Fusar‐Poli, L., Freeman, R. E., Spreen, M., Ket, J. C., Vink, A. C., ... & Gold, C. (2017). Music therapy for depression. Cochrane database of systematic reviews, (11).

ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot(), to correct syntax errors, and for first draft of abstract. Accessed `r Sys.Date()`.

Clauw, Daniel J. “Fibromyalgia: An Overview.” The American Journal of Medicine, vol. 122, no. 12, Dec. 2009, doi:10.1016/j.amjmed.2009.09.006. 

Eduardo A. Garza-Villarreal, Victor Pando-Naude, Fernando A. Barrios, Peter Vuust, Lene Vase, Elvira Brattico, Erick H. Pasaye, and Sarael Alcauter (2019). Functional Connectivity of Music-Induced Analgesia in Fibromyalgia. OpenNeuro. [Dataset] doi: 10.18112/openneuro.ds001928.v1.1.0