---
title: "Music and Chronic Pain"
author: "Norah Saarman"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# ABSTRACT
This study investigates whether listening to music can reduce the sensation of pain in individuals diagnosed with chronic pain due to Fibromyalgia. Participants rated their pain before and after listening to either five minutes of self-selected pleasant music or a pink-noise control stimulus. Boxplot visualizations and paired t-tests were used to compare pre- and post-treatment pain levels across both conditions. Results showed a statistically significant decrease in pain following the music condition, while the control condition produced no meaningful reduction and even suggested a slight increase in reported pain. These findings support the hypothesis that music can serve as an effective non-pharmacological method for reducing chronic pain. Future research with larger samples and broader auditory conditions may further clarify the mechanisms underlying music-induced analgesia.

# BACKGROUND

Music has been a deep and important part of human culture and expression for millennia. Yet, not much is understood about how music affects the human body physiologically. As research develops, music has been shown to provide a myriad of benefits to human quality of life. Recently, S. Aalbers and others found that when combined with treatment as usual, music had a significant positive effect on patients with depression (Aalbers, 2017).

This study explores how music affects the pain sensation of individuals diagnosed with chronic pain. The results will provide exciting insights to a potential link between music and pain reduction.

# STUDY QUESTION and HYPOTHESIS

## Questions  

Does music affect the sensation of pain in individuals with chronic pain?

## Hypothesis 

Listening to music decreases the sensation of pain in individuals with chronic pain.

## Prediction

Participants who listen to music will report a lower sensation of pain.

# METHODS

The dataset was provided by Pando-Naude et al. (5). The study involved 20 individuals diagnosed with Fibromyalgia (FM). The original study had participants rate their pain level on a scale of 1-10 before and after listening to 5 minutes of either music chosen by the individual or pink noise (control). By requirement, the music had a tempo less than 120 bpm and was deemed familiar and pleasant by the participant.

##*Figure 1* - Boxplot of Before and After Control and Treatment
Visualization of 4 data groups (Before Control, After Control, Before Music, After Music) was presented via boxplots. All 4 plots are presented first, then control and treatment groups are separated for clarity.

```{r total boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#control treatment
data_total <- data %>%
  select(1, 5, 6, 7, 8) %>%
  pivot_longer(
    cols = c(2, 3, 4, 5),
    names_to = "treatment",
    values_to = "value" )


#plot control
    ggplot(data_total, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_manual(values = c("pic1" ="#B16C02" , "pic2" = "#FFB501", "pim1" ="#003F80" , "pim2" = "#1E88E5")) +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
      scale_x_discrete(labels = c("pic1" = "Before Control", "pic2" = "After Control", "pim1" = "Before Music", "pim2" = "After Music"))
    
```

##*Figure 2* - Control Boxplots: Before and After
```{r control boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#control treatment
data_control <- data %>%
  select(1, 5, 6) %>%
  pivot_longer(
    cols = c(2, 3),
    names_to = "treatment",
    values_to = "value" )


#plot control
    ggplot(data_control, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
  
    theme_classic() +
    theme(legend.position="none") +
    scale_fill_manual(values = c("pic1" ="#B16C02" , "pic2" = "#FFB501")) +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
      scale_x_discrete(labels = c("pic1" = "Before Control", "pic2" = "After Control"))
```


##*Figure 3* - Treatment Boxplots: Before Music and After Music
```{r music boxplot, warning=FALSE}
library(ggplot2)
library(tidyr)
library(dplyr)

#changed column setup to make boxplot work
data <- read.csv("musicdataset.csv")
#music treatment
data_music <- data %>%
  select(1, 7, 8) %>%
  pivot_longer(
    cols = c(2, 3),
    names_to = "treatment",
    values_to = "value")

#plot music
    ggplot(data_music, aes(x=treatment, y=value, fill=treatment)) + 
    geom_boxplot(alpha=0.6) +
    
    scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, 1)) +
    scale_fill_manual(values = c("pim1" ="#003F80" , "pim2" = "#1E88E5")) +
    theme_classic() +
    theme(legend.position="none") +
    theme(axis.text.x = element_text(size=12)) +
      theme(axis.title.x = element_text(size=16, face="bold", margin = margin(t = 10))) +
    theme(axis.text.y = element_text(size=12)) + 
    theme(axis.title.y = element_text(size=16, face="bold", margin = margin(r = 10))) +
    theme(plot.title = element_text(face = "bold", size=16, margin = margin(b = 10))) +
    ylab("Pain Level") +
    xlab("") +
    scale_x_discrete(labels = c("pim1" = "Before Music", "pim2" = "After Music"))
    
```



##Statistical Analysis
T-test was performed to analyze mean between before and after control and test fro significance.
```{r t test control}
t.test(data$pic2, data$pic1, paired = TRUE)

```
T-test was performed to analyze mean between before and after music and test for significance.
```{r t test music}
t.test(data$pim2, data$pim1, paired = TRUE)

```

# DISCUSSION
The results show statistical evidence suggesting that music leads to a greater decrease in pain after subjects listened to music. The control exhibited moderate correlation for the pink noise to affect pain levels in the positive direction, meaning that there may be evidence suggesting that pink noise increased pain. The treatment group showed a strong statistical significance, rejecting the null hypothesis that music does not affect pain sensation. Furthermore, there was a greater mean difference that was in the negative direction for music, reflecting a visible decrease in the average reported pain levels in patients.

The major limitation was each person's interpretation of pain because of each individual's personal experiences and pain tolerances. A way to reduce the effects of this is to use a visual pain scale, such as the “Wong-Baker FACES Scale,” that can provide a visual reference point to help patients determine pain levels on a scale that could be more universal. The experiment utilized data from only 20 patients. Future research may benefit from a larger sample size in order to achieve more concise data.

Further research should compare both different forms of “noise” (white, brown, etc.) and multiple music genres to other methods of pain reduction, such as medication or physical/talk therapy. Regardless, because of the idiopathic cause of this disease, the evidence showing decreased pain from music is a step towards a more comfortable future in patients who have this diagnosis, as they may find affordable and effective methods for finding relief.

# CONCLUSION
After testing the pain sensation of individuals with chronic pain due to Fibromyalgia with and without music, it can be concluded with strong statistical significance that listening to familiar, pleasant music reduces the sensation of pain.

# REFERENCES     
Aalbers, S., Fusar‐Poli, L., Freeman, R. E., Spreen, M., Ket, J. C., Vink, A. C., ... & Gold, C. (2017). Music therapy for depression. Cochrane database of systematic reviews, (11).

ChatGPT. OpenAI, version Jan 2025. Used as a reference for functions such as plot(), to correct syntax errors, and for first draft of abstract. Accessed `r Sys.Date()`.

Clauw, Daniel J. “Fibromyalgia: An Overview.” The American Journal of Medicine, vol. 122, no. 12, Dec. 2009, doi:10.1016/j.amjmed.2009.09.006. 

Eduardo A. Garza-Villarreal, Victor Pando-Naude, Fernando A. Barrios, Peter Vuust, Lene Vase, Elvira Brattico, Erick H. Pasaye, and Sarael Alcauter (2019). Functional Connectivity of Music-Induced Analgesia in Fibromyalgia. OpenNeuro. [Dataset] doi: 10.18112/openneuro.ds001928.v1.1.0